---
title: "corr ~ predicted IC content buccal samples at baseline"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F, message = F, warning = F, eval = F)
```


```{r libs}
library(tidyverse)

source("src/corrIcBaseline.R")
```

This script is to compute the correlation with predicted IC buccal samples (epigenetics) at baseline for all saliva microbiome features.

# Compute correlation at baseline for normalized data

Normalized data (!) are loaded, meaning z-scaled relative abundances for microbiome features, and z-scaled estimated IC content buccals

```{r getdata, eval = F}

# load in data
load("data/data_normalized.Rdata")

exp <- c("Saliva microbiome: ASVs",
         "Saliva microbiome: families")

# extract all data in long format plus basic anno
df <- as.data.frame(longFormat(data[,,exp], colData = c("interventionId", "visitId"))) |> 
  dplyr::filter(!is.na(value)) %>%
  dplyr::filter(visitId == "M0" & interventionId != "S") %>%
  dplyr::select(-interventionId)

# grab IC content buccals
exp <- c("Composite methylation scores: buccal")
features <- c("interventionId", "visitId")
dat <- as.data.frame(wideFormat(data[,,exp],
                                   colData = features)) %>% 
  dplyr::filter(visitId == "M0" & interventionId != "S") %>%
  dplyr::select(-interventionId)
colnames(dat) <- gsub("Composite.methylation.scores..buccal_", "", colnames(dat))
dat <- dat %>%
  dplyr::select(primary,ic)

df <- full_join(df,dat) 

rm(data,dat) ; gc()

corr <- corrIcBaseline(df)

save(corr, file = "out/corrIcBaseline_saliva.R")

```


# Compute correlation at baseline for raw data

Raw data (!) are loaded, meaning relative abundances for microbiome features, and estimated IC content buccals

```{r corr, eval = F}

# load in data
load("data/data_raw.Rdata")

exp <- c("Saliva microbiome: ASVs",
         "Saliva microbiome: families")

# extract all data in long format plus basic anno
df <- as.data.frame(longFormat(data[,,exp], colData = c("interventionId", "visitId"))) |> 
  dplyr::filter(!is.na(value)) %>%
  dplyr::filter(visitId == "M0" & interventionId != "S") %>%
  dplyr::select(-interventionId)

# grab IC content buccals
exp <- c("Composite methylation scores: buccal")
features <- c("interventionId", "visitId")
dat <- as.data.frame(wideFormat(data[,,exp],
                                   colData = features)) %>% 
  dplyr::filter(visitId == "M0" & interventionId != "S") %>%
  dplyr::select(-interventionId)
colnames(dat) <- gsub("Composite.methylation.scores..buccal_", "", colnames(dat))
dat <- dat %>%
  dplyr::select(primary,ic)

df <- full_join(df,dat) 

rm(data,dat) ; gc()

corr <- corrIcBaseline(df)

save(corr, file = "out/corrIcBaseline_saliva_raw.R")

```



