---
title: "Variance decomposition"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```


```{r libs}
library(dplyr)
library(here)
library(ggplot2)
library(MultiAssayExperiment)
library(lme4)
here::i_am("helper-variance-decomposition.qmd")
```

Here, we partition variance of omic features.

## Preparing the data

First, we load and prepare the normalized data:

```{r}
load("data/data_normalized.Rdata")

# Removing features with zero variance:
for (i in 1:length(experiments(data))){
  experiments(data)[[i]] <- experiments(data)[[i]][apply(assay(data[[i]]), 1, function(x) var(x,na.rm = T) != 0), ]
}

# Extract relevant omic and covariate data:
omes <- c(1, 3, 4, 5, 6, 9, 10, 13:16, 22, 28, 29)

df <- longFormat(data[,data$interventionId != "S" & !data$visitId %in% c("M12", "M18"),omes],
                 colDataCols=c("interventionId", "subjectId", "visitId", "age_at_consent", "compliance", "bmi_at_consent")) |> 
  as.data.frame()

# Removing superfluous variables (non-utilized composite scores, immune populations)
indices_cerv <- readxl::read_xlsx(here("src/indices.xlsx"), sheet = 1)
indices_buccal <- readxl::read_xlsx(here("src/indices.xlsx"), sheet = 2)
indices_bl <- readxl::read_xlsx(here("src/indices.xlsx"), sheet = 3)
load(here("src/populations_names_annotated.Rdata"))
populations <- populations |> 
  dplyr::filter(staining != 't cell stimulation' & main_analysis == 'yes')

df <- df |> 
  dplyr::filter(!grepl("Composite", assay) | (grepl("Composite", assay) & grepl('cervical', assay) & rowname %in% indices_cerv$x) |
                  (grepl("Composite", assay) & grepl('buccal', assay) & rowname %in% indices_buccal$x) |
                  (grepl("Composite", assay) & grepl('blood', assay) & rowname %in% indices_bl$x)) |> 
  dplyr::filter(!grepl("Flow cytometry", assay) | (grepl("Flow cytometry", assay) & rowname %in% populations$name))

rm(data);gc()
```

Tidy up factor data:

```{r}
df$visitId <- factor(as.character(df$visitId), levels = c("M0", "M2", "M4", "M6"))
df$compliance <- factor(df$compliance, levels = c("low", "medium", "high"))
df$interventionId <- factor(df$interventionId, levels = c("I", "K"))
df$subjectId <- factor(df$subjectId)
df <- df[!is.na(df$value),]
```

Code unique feature ids by concatenating the assay and feature name:

```{r}
# feature IDs
df$featureid <- paste0(df$assay,"_",df$rowname)
featureid <- unique(df$featureid)
```

Counting the number of features and number of unique subject to filter featureids that have too few observations:

```{r}
counts <- df |> 
  as.data.frame() |> 
  dplyr::group_by(featureid) |> 
  dplyr::reframe(n = dplyr::n(),
                 n_id = length(unique(subjectId))) |> 
  dplyr::ungroup() |> 
  dplyr::mutate(ratio = n/n_id) |> 
  dplyr::filter(ratio > 1.2)
featureid <- featureid[featureid %in% counts$featureid]
```

## LMMs for variance decomposition

The below code is parallelized and takes approximately ~4s for 400 features.

```{r}
out.lmm <- parallel::mclapply(featureid, function(i) {
    tryCatch(lmer(value ~ visitId*compliance + age_at_consent + bmi_at_consent + interventionId + (1 | subjectId),
                  data = as.data.frame(df[df$featureid==i,]),
                  REML = F),
             error = function(e) return(e))
}, mc.preschedule = T, mc.cores = 8)
names(out.lmm) <- featureid
save(out.lmm, file = "out/variance_partition_lmm.Rdata") # save intermediate out.lmm (optional)
```


## Extracting the variance from models

```{r}
# Extract variance
df.var <- dplyr::bind_rows(lapply(out.lmm, function(x) as.data.frame(insight::get_variance(x))), .id="featureid")
df.var$f.var.fixed <- df.var$var.fixed/(df.var$var.fixed + df.var$var.random + df.var$var.residual)
df.var$f.var.random <- df.var$var.random/(df.var$var.fixed + df.var$var.random + df.var$var.residual)
df.var$f.var.residual <- df.var$var.residual/(df.var$var.fixed + df.var$var.random + df.var$var.residual)
# Assay group
df.var <- df.var |>
  # reframe
  dplyr::select(featureid, starts_with("f.var")) |>
  dplyr::mutate(assay = stringr::str_split(featureid, "_", simplify = T)[,1]) |>
  dplyr::rowwise() |> 
  dplyr::mutate(icc = f.var.random/sum(f.var.random+f.var.fixed+f.var.residual)) |> 
  dplyr::ungroup()
save(df.var, file = "out/variance_partition_df_var.Rdata")
```

ICC is computed as V ran/V total and visualised for each ome.