---
title: "Major gut microbiota-derived metabolites"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, eval = F)
```


```{r libs}
library(tidyverse)
library(kableExtra)
library(openxlsx)
```

Based on review article Yang and Cong 2021, Cellular & Molecular Immunology (https://doi.org/10.1038/s41423-021-00661-4).

# Overview

1. Intermittent fasting alters gut microbiota, functional predictions in relation to gut microbiota-derived metabolites? <br> 

2. Further evidence of increased excretion of these metabolites in urine? <br>

3. Modulation in blood immune cells? <br>

# Scan functional microbiome data for key enzymes in selected pathways

```{r}
df <- data.frame(metabolite.class = c(rep("SCFA",9)),
                 Synthesis = c("Butyrate",
                             "Butyrate",
                             "Butyrate",
                             "Propionate",
                             "Propionate",
                             "Propionate",
                             "Propionate",
                             "Formate",
                             "Acetate"),
                 enzyme = c("EC:2.8.3.8",
                            "EC:2.8.3.9",
                            "EC:2.7.2.7",
                            "EC:2.8.3.18",
                            "EC:2.8.3.1",
                            "EC:2.7.2.15",
                            "EC:2.7.2.15",
                            "EC:2.3.1.54",
                            "EC:2.7.2.1"),
                 Knumber = c("K01034",
                             "K01035",
                             "K00929",
                             "K18118",
                             "K01026",
                             "K00932",
                             "K19697",
                             "K00656",
                             "K00925"),
                 name = c("acetate CoA/acetoacetate \n CoA-transferase alpha subunit",
                          "acetate CoA/acetoacetate \n CoA-transferase beta subunit",
                          "butyrate kinase",
                          "succinyl-CoA:acetate CoA-transferase",
                          "propionate CoA-transferase",
                          "propionate kinase",
                          "propionate kinase",
                          "formate C-acetyltransferase",
                          "acetate kinase"))

df2 <- data.frame(metabolite.class = c(rep("Tryptophan catabolism",4)),
                 Synthesis = c("Indole",
                             "Tryptophan",
                             "Indolepyruvate",
                             "Indole-3-acetamide"),
                 enzyme = c("EC:4.1.99.1",
                            "EC:4.1.1.105",
                            "EC:2.6.1.27",
                            "EC:1.13.12.3"),
                 Knumber = c("K01667",
                             "K01593",
                             "K14265",
                             "K00466"),
                 name = c("tryptophanase",
                          "L-tryptophan decarboxylase",
                          "tryptophan aminotransferase",
                          "tryptophan 2-monooxygenase"))

df <- full_join(df,df2)

df2 <- data.frame(metabolite.class = c(rep("Trimethylamine",3)),
                 Synthesis = c("L-carnitine reduction",
                             "L-carnitine reduction",
                             "Choline degradation"),
                 enzyme = c("EC:1.14.13.239",
                            "EC:1.14.13.239",
                            "EC:4.3.99.4"),
                 Knumber = c("K22443",
                             "K22444",
                             "K20038"),
                 name = c("carnitine monooxygenase",
                          "carnitine monooxygenase",
                          "choline trimethylamine-lyase"))

df <- full_join(df,df2)

df2 <- data.frame(metabolite.class = c(rep("Secondary bile acids",1)),
                 Synthesis = c("Deoxycholate/Chenodeoxycholate"),
                 enzyme = c("EC:1.1.1.159"),
                 Knumber = c("K00076"),
                 name = c("7-alpha-hydroxysteroid dehydrogenase"))

df <- full_join(df,df2)

df2 <- data.frame(metabolite.class = c(rep("BCAA",6)),
                 Synthesis = c("L-isoleucine",
                             "L-valine",
                             "L-valine",
                             "L-valine",
                             "L-leucine",
                             "L-BCAA"),
                 enzyme = c("EC:4.3.1.19",
                            "EC:2.2.1.6",
                            "EC:2.2.1.6",
                            "EC:2.2.1.6",
                            "EC:4.2.1.33",
                            "EC:2.6.1.42"),
                 Knumber = c("K17989",
                             "K01652",
                             "K01653",
                             "K11258",
                             "K01702",
                             "K00826"),
                 name = c("L-threonine dehydratase/deaminase",
                          "acetolactate synthase",
                          "acetolactate synthase",
                          "acetolactate synthase",
                          "isopropylmalate isomerase",
                          "branched-chain aminotransferase"))

df <- full_join(df,df2)

df2 <- data.frame(metabolite.class = c(rep("Vitamins",11)),
                 Synthesis = c("Vitamine K2",
                             "Vitamine K2",
                             "Vitamine K2",
                             "Vitamine B1",
                             "Vitamine B2",
                             "Vitamine B3",
                             "Vitamine B5",
                             "Vitamine B6",
                             "Vitamine B7",
                             "Vitamine B9",
                             "Vitamine B12"),
                 enzyme = c("EC:2.5.1.74",
                            "EC:2.1.1.163",
                            "EC:1.21.98.1",
                            "EC:3.1.3.1",
                            "EC:2.5.1.9",
                            "EC:6.3.4.21",
                            "EC:6.3.2.1",
                            "EC:2.6.99.2",
                            "EC:2.8.1.6",
                            "EC:2.5.1.15",
                            "EC:1.3.1.54"),
                 Knumber = c("K02548",
                             "K03183",
                             "K11784",
                             "K01077",
                             "K00793",
                             "K00763",
                             "K01918",
                             "K03474",
                             "K01012",
                             "K00796",
                             "K05895"),
                 name = c("1,4-dihydroxy-2-naphthoate polyprenyltransferase",
                          "demethylmenaquinone methyltransferase",
                          "cyclic dehypoxanthinyl futalosine synthase",
                          "alkaline phosphatase",
                          "riboflavin synthase",
                          "nicotinate phosphoribosyltransferase",
                          "pantoate--beta-alanine ligase",
                          "pyridoxine 5-phosphate synthase",
                          "biotin synthase",
                          "dihydropteroate synthase",
                          "precorrin-6A/cobalt-precorrin-6A reductase"))

df <- full_join(df,df2)

kable(df)

#saveRDS(df, file = "out/gut_metabolites_functional_scan1.Rds")

```




