---
title: "Permutation strategy validation under the null hypothesis"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---
  
```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, eval = F)
```


```{r libs, eval = T}

library(dplyr)
library(data.table)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(purrr)
library(tidyr)
library(ggplot2)
library(future.apply)

here::i_am("perm_validation.qmd")

```

This analysis evaluates whether the permutation procedure controls Type I error under the null hypothesis.

The simulated data contain:
  
- subject-level random effects,
- repeated measurements,
- intervention groups,
- age and BMI covariates,

but **no visit effect and no visit × intervention interaction**.

A valid permutation procedure should generate approximately uniform p-values and maintain a false-positive rate close to the nominal alpha level (0.05).


```{r design, eval = F}
# load in data
load("data/data_normalized.Rdata")

load("src/vars.Rdata")

varsList <- vars |> dplyr::filter(!assay %in% c('Saliva microbiome: ASVs', 'Stool microbiome: ASVs',
                                            'Saliva microbiome: families_clr', 'Stool microbiome: families_clr')) |> 
  dplyr::group_by(assay) |> 
  dplyr::summarise(features = list(x), .groups = 'drop') |> tibble::deframe()

# extract all data in long format plus basic anno
df <- as.data.frame(longForm(subsetByAssay(data,varsList),
                             colData = c("age_at_consent", "interventionId", "subjectId", "visitId", "time",
                                                 'compliance', 'comprate', 'supplement',
                                                 'bmi_at_consent',
                                                 'dbmi',
                                                 'mpstatrs'))) |> 
  dplyr::filter(!is.na(value))


# remove S and any non M0-M6
df <- df |>
  dplyr::filter(interventionId!='S' & !visitId %in% c("M12", "M18")) |>
  dplyr::mutate(compliance = factor(compliance, levels = c("low", 'medium', 'high')),
                visitId = factor(visitId, levels = c("M0", "M2", "M4", "M6")),
                visitIdOrdered = ordered(visitId)) |> 
  dplyr::mutate(rowname = paste0(assay, "_", rowname))

# check class
class(df$visitId)
class(df$visitIdOrdered)

# remove full data
rm(data);gc()

design <- df %>% dplyr::select(subjectId, interventionId, visitId, age_at_consent, bmi_at_consent) %>% distinct()
```

```{r sim_par}

n_null_simulations <- 100
n_permutations <- 100
alpha <- 0.05
set.seed(134)

```

```{r functions}

#value = age + bmi + subject random effect + residual
#no visit effect

simulate_null_dataset <- function(design){

  subject_effects <- data.frame(
    subjectId = unique(design$subjectId),
    subject_random = rnorm(
      length(unique(design$subjectId)),
      mean = 0,
      sd = 1
    )
  )

  simulated <- design %>%
    left_join(subject_effects, by="subjectId") %>%
    mutate(
      value =
        0.05 * age_at_consent +
        0.02 * bmi_at_consent +
        subject_random +
        rnorm(n(),0,1)
    )

  return(simulated)

}

# reproduce permutation strategy
# shuffle visitId within subject, preserve longitudinal structure

run_one_permutation <- function(df, test_term){

  df_perm <- copy(df)

  setDT(df_perm)

  df_perm[, visitId :=
           if(.N > 1)
             sample(visitId)
           else
             visitId,
         by = subjectId]


  model <- lmer(
    value ~ visitId*interventionId +
      age_at_consent +
      bmi_at_consent +
      (1|subjectId),
    data=df_perm,
    REML=FALSE
  )


  result <- broom.mixed::tidy(model) %>%
    filter(
      effect=="fixed",
      grepl(paste0(test_term,"$"),term)
    ) %>%
    pull(p.value)


  return(result)

}


run_permutation_analysis <- function(df, test_term){

  # observed model

  observed_model <- lmer(
    value ~ visitId*interventionId +
      age_at_consent +
      bmi_at_consent +
      (1|subjectId),
    data=df,
    REML=FALSE
  )


  observed_p <- broom.mixed::tidy(observed_model) %>%
    filter(
      effect=="fixed",
      grepl(paste0(test_term,"$"),term)
    ) %>%
    pull(p.value)


  # permutations

  permuted_p <- map_dbl(
    1:n_permutations,
    ~run_one_permutation(df, test_term)
  )


  empirical_p <-
    mean(permuted_p <= observed_p)


  tibble(
    observed_p = observed_p,
    empirical_p = empirical_p,
    permuted_distribution = list(permuted_p)
  )

}


```

# Simulation visitID

`r n_null_simulations` simulations, `r n_permutations` permutations for visitId = M6.


```{r run_m6, eval = F}

testing <- "visitIdM6"

null_results <- map_dfr(
  1:n_null_simulations,
  function(i){

    message("Simulation ",i)

    simulated <- simulate_null_dataset(design)

    run_permutation_analysis(simulated, testing)

  }
)

save(null_results, file = "out/perm_sim_m6.Rdata")

```

## Type I error estimation

```{r}

load("out/perm_sim_m6.Rdata")

type1_error <- mean(
  null_results$empirical_p < alpha
)

type1_error

tibble(
  expected = alpha,
  observed = type1_error
)

```

## Distribution of permutation p-values

### Across null simulations

Under the null, permutation p-values should approximately follow a uniform distribution (P value exact one-sample Kolmogorov-Smironov test > 0.05).

```{r}

p1 <- null_results %>%
  ggplot(
    aes(empirical_p)
  )+
  geom_histogram(
    bins=20
  )+
  xlim(0,1)+
  labs(
    x="Empirical permutation P value",
    y="Count",
    title="Term = visitIdM6"
  ) +
  geom_vline(xintercept = 0.05, linetype = "dashed") +
  theme_classic()


ks_test <- broom::tidy(ks.test(
  null_results$empirical_p,
  "punif"
))

ks_test

summary_m6 <- tibble(
  Term = testing,
  Number_of_simulations = n_null_simulations,
  Number_of_permutations = n_permutations,
  Nominal_alpha = alpha,
  Observed_Type_I_error = type1_error,
  KS_uniformity_pvalue = ks_test$p.value
)


```

### Underlying p-values (pooled)

```{r}
all_perm_p <- unlist(null_results$permuted_distribution)
ggplot(data.frame(p=all_perm_p), aes(p)) +
  geom_histogram(bins=20) +
  xlim(0,1) +
  labs(
    x="P values generated during permutation under the null",
    y="Count",
    title="Term = visitIdM6"
  ) +
  theme_classic()

ks.test(all_perm_p, "punif")
```

# Simulation interaction

`r n_null_simulations` simulations, `r n_permutations` permutations for visitIdM6:interventionIdK.


```{r run_m6, eval = F}

testing <- "visitIdM6:interventionIdK"

null_results <- map_dfr(
  1:n_null_simulations,
  function(i){

    message("Simulation ",i)

    simulated <- simulate_null_dataset(design)

    run_permutation_analysis(simulated, testing)

  }
)

save(null_results, file = "out/perm_sim_int.Rdata")

```

## Type I error estimation

```{r}

load("out/perm_sim_int.Rdata")

type1_error <- mean(
  null_results$empirical_p < alpha
)

type1_error

tibble(
  expected = alpha,
  observed = type1_error
)

```

## Distribution of permutation p-values

### Across null simulations

Under the null, permutation p-values should approximately follow a uniform distribution (P value exact one-sample Kolmogorov-Smironov test > 0.05).

```{r}

p2 <- null_results %>%
  ggplot(
    aes(empirical_p)
  )+
  geom_histogram(
    bins=20
  )+
  xlim(0,1)+
  labs(
    x="Empirical permutation P value",
    y="Count",
    title="Term = visitIdM6:interventionIdK"
  ) +
  geom_vline(xintercept = 0.05, linetype = "dashed") +
  theme_classic()


ks_test <- broom::tidy(ks.test(
  null_results$empirical_p,
  "punif"
))

ks_test

summary_int <- tibble(
  Term = testing,
  Number_of_simulations = n_null_simulations,
  Number_of_permutations = n_permutations,
  Nominal_alpha = alpha,
  Observed_Type_I_error = type1_error,
  KS_uniformity_pvalue = ks_test$p.value
)


```

### Underlying p-values (pooled)

```{r}
all_perm_p <- unlist(null_results$permuted_distribution)
ggplot(data.frame(p=all_perm_p), aes(p)) +
  geom_histogram(bins=20) +
  xlim(0,1) +
  labs(
    x="P values generated during permutation under the null",
    y="Count",
    title="Term = visitIdM6"
  ) +
  theme_classic()

ks.test(all_perm_p, "punif")
```


# Summary

```{r}

summary_table <- bind_rows(summary_m6, summary_int)
summary_table

```

```{r print.supplementary, eval = F}

p <- p1 + p2

ggsave(p,
       filename = 'figures-png/val_perm.png',
       dpi = 600,
       width = 10,
       height = 4,
       unit = 'in')
```


